```{r}
#| echo: false
source("_common.R")
```


# Mixed models (L11-1) {.unnumbered}

```{r}
#| echo: false
#| output: asis
lo_dir <- "../BIO144_Course_Information/learning_objectives"
source(file.path(lo_dir, "lo.R"))
lo_callout("11.1", lo_dir)
```

::: {.callout-note icon=false}
## Code walkthrough for this chapter
<!-- 2027 addition: code walkthrough script and video -->

**Script:** [walkthrough_ch11_mixed_models.R](walkthroughs/walkthrough_ch11_mixed_models.R){download="walkthrough_ch11_mixed_models.R"}

This R script contains the code of this chapter, with comments that explain each step. Download it, save it in your BIO144 RStudio project folder, open it in RStudio, and run it line by line while you read the chapter or watch the video. The script reads the datasets directly from the internet. Its last section ("Practical toolbox") shows code that the Unit 11 practical uses but this chapter does not.

To download the script straight into your project folder (for example on RStudio Server), run this line in the Console:

`download.file("https://opetchey.github.io/BIO144_Course_Book/walkthroughs/walkthrough_ch11_mixed_models.R", destfile = "walkthrough_ch11_mixed_models.R")`

**Video walkthrough:** coming soon. <!-- VIDEO PLACEHOLDER: replace "coming soon." with the link, e.g. [Watch the video walkthrough](https://youtu.be/XXXX), or embed it with the Quarto video shortcode -->
:::

## Introduction

So far in BIO144 we mostly used models where **each observation is assumed independent**:

* linear models (`lm()`)
* generalized linear models (`glm()`), e.g. Poisson and binomial.

By independent we mean that the value of one observation does not give us any information about the value of another observation.

But many biological datasets violate independence because observations come in **groups**:

- repeated measures on the same individual (before/after, time series, multiple tissues)
- multiple individuals from the same plot / site / stream / lake / cage / family
- students within classes, patients within hospitals, samples within batches

In these cases, treating all rows as independent often leads to **false confidence** (too-small standard errors, too-small p-values). Mixed models are one standard way to handle this.

::: {.callout-important}
**Core idea** A *mixed model* extends over all our previous models by adding **random effects** that represent grouping structure (clusters) in the data.

* **fixed effects**: effects you want to estimate explicitly (treatments, temperature, time, …)
* **random effects**: variation among groups (individuals, sites, years, …) that induces correlation within groups
:::

<!---
**Think–Pair–Share** (#tps-why-mixed) Name one example in biology where multiple measurements come from the same "unit" (individual, plot, lake, strain, batch…). What goes wrong if we pretend those measurements are independent?
--->

## Why not just average?

A common workaround is to average repeated measurements to a single value per group and then use `lm()`.

Sometimes this is OK — but often it throws away information.

Reasons *not* to average include:

1. **Imbalanced sampling** (groups have different numbers of observations): averaging changes the weighting.
2. **Which average?** (mean, median, mode): different choices answer different questions.
3. **False confidence**: pretending "n rows" are independent can make uncertainty look too small.
4. **You may want to study variation among groups** (some individuals respond more strongly than others).
5. **“Sharing information” across groups**: mixed models partially pool group estimates toward the overall mean.
6. **Keeping information**: you can use all observations without collapsing the design.

::: {.callout-note}
A mixed model gives you a principled compromise between:

* analysing each observation as a independent one, and
* averaging everything (too coarse).

This is sometimes called **partial pooling**.
:::

## The problem: non-independence and pseudoreplication

Imagine we measure the same individual multiple times.

If we fit a simple linear model, the residuals are assumed independent:

$$\varepsilon_i \sim \text{Normal}(0, \sigma^2), \quad \text{independent across } i$$

But repeated measures induce correlation:

- measurements from the same individual tend to be more similar
- so residuals are not independent

This is one common form of **pseudoreplication**: treating repeated measurements as if they were separate independent replicates.

## Random intercept models

### Random intercept idea

Suppose we measure a response $y$ (e.g. growth) for individuals $j$ at observations $i$. And that we measure each individual more than once. And that we have some treatment, such as temperature $x$.

This means that our data have a **grouping structure**: observations are grouped by individual. If we want to calculate the mean growth, we should account for this grouping.

```{r}
#| echo: false
set.seed(42)
n_individuals <- 10
n_obs_per_ind <- 5
individual_ids <- factor(sprintf("Ind%02d", 1:n_individuals))
example_repeated <- expand_grid(
  individual = individual_ids,
  obs_number = 1:n_obs_per_ind) |> 
  mutate(x = runif(n_individuals * n_obs_per_ind, 0, 10)
) |>
  arrange(individual) |>
  mutate(
    # random intercept per individual
    b0 = rnorm(n_individuals, 0, 2)[as.integer(individual)],
    # fixed effects
    beta0 = 5,
    beta1 = 1.5,
    # response variable
    y = beta0 + beta1 * x + b0 + rnorm(n(), 0, 1)
  ) |>
  select(individual, temperature = x, growth = y)
# 2027 fix: relative path instead of here()
write_csv(example_repeated, "datasets/mixed_model_example.csv")
```

Let's read in the illustrative data:

```{r}
example_repeated <- read_csv("datasets/mixed_model_example.csv")
```

Here are the first few rows:

```{r}
head(example_repeated)
```

In this data we have two sources of variation in `growth`:

1. **fixed effect** of `temperature` (same for all individuals)
2. **random effect** of `individual` (different baseline growth for each individual)

A fixed effect is something we want to estimate explicitly (e.g. how growth changes with temperature). Fixed effects are variables in which the values have specific meaning (e.g. temperature = 5°C).

A random effect is something that varies among groups (individuals here), but where we are not interested in the specific values for each group. Instead, we want to estimate the amount of variation among groups. Random effects are variables where the specific values are not of interest, but rather the variation among them (e.g. individual identity). I.e., the value "Ind01" has no specific meaning; we just want to know how much individuals differ from each other on average. That value could be anything.

A random effect is often containing levels that are considered a random sample from a larger population. We did not, for example, choose specific individuals for a reason; they are just a sample of all possible individuals.

An appropriate model for this data is a **random-intercept mixed model**. It is a model in which each group (individual) has its own intercept (baseline), but the intercepts are assumed to come from a common distribution.

A random-intercept mixed model is:

$$y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0j} + \varepsilon_{ij}$$

where:

- $\beta_0, \beta_1$ are **fixed effects**
- $x_{ij}$ is the explanatory variable for observation $i$ in group $j$. Also called a **fixed effect**.
- $b_{0j}$ is a **random intercept** for group $j$, typically $b_{0j} \sim \text{Normal}(0,\sigma_b^2)$
- $\varepsilon_{ij}$ is residual error, $\varepsilon_{ij} \sim \text{Normal}(0,\sigma^2)$

Interpretation:

- each group gets its own intercept (baseline), but
- those intercepts are assumed to come from a common distribution

### Random intercept syntax in R

To make a mixed model in R we can no longer use `lm()` or `glm()`. Instead, we use the `lmer()` function from the `lme4` package.

In the `lmer` function we have to specify the random effects in a special way. For example, a random-intercept model for `y` with fixed effect `x` and random intercept by `group` is specified as:

```r
y ~ x + (1 | group)
```

Read it as: "a model for `y` with fixed effect `x` and a random intercept by `group`".

## Random slope models

Sometimes groups differ not only in baseline level, but also in how they respond to an explanatory variable.

A random-slope model:

$$y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0j} + b_{1j}x_{ij} + \varepsilon_{ij}$$

In R:

```r
y ~ x + (1 + x | group)
```

Here each group has its own intercept **and** its own slope, and these can be correlated.

::: {.callout-caution}
Random slopes are powerful but can be hard to estimate with small datasets.
If the model struggles to fit (singular fit warnings), consider simplifying.
:::

## Nested and crossed random effects

### Nested

**Nested** means one grouping factor is contained within another.

Example: measurements within plants within plots:

```r
y ~ treatment + (1 | plot/plant)
```

This expands to `(1 | plot) + (1 | plot:plant)`.

### Crossed

**Crossed** means groups are not nested.

Example: repeated measures with multiple observers (each observer measures many individuals; each individual is measured by many observers):

```r
y ~ x + (1 | individual) + (1 | observer)
```

## Hands-on example: plant growth with repeated measures

### Biological story

You are studying how temperature affects plant growth.

- 30 plants are grown at one of three temperatures: 10°C, 15°C, 20°C.
- Each plant is measured weekly for 8 weeks.
- Response: plant height (cm).

Because we repeatedly measure the **same plant**, the observations are not independent.
We will compare:

1. a naive linear model (wrong independence assumption),
2. a mixed model with plant as a random effect.

### An example dataset

```{r}
#| echo: false
set.seed(11)

n_plants <- 30
weeks <- 0:7

plant_tbl <- tibble(
  plant_id = factor(sprintf("P%02d", 1:n_plants)),
  temp = factor(rep(c("10", "15", "20"), length.out = n_plants))
)

# Random intercept per plant (baseline differences)
plant_intercepts <- rnorm(n_plants, mean = 0, sd = 1.5)

# Fixed effects
beta0 <- 10        # baseline height at week 0 (cm)
beta_week <- 1.8   # growth per week
beta_temp <- c("10" = -1.0, "15" = 0.0, "20" = 1.2)  # temp effect on baseline

# Optional: allow temp to affect growth rate a little
beta_week_temp <- c("10" = -0.2, "15" = 0.0, "20" = 0.25)

plant_growth <- expand_grid(plant_tbl, week = weeks) |>
  arrange(plant_id, week) |>
  mutate(
    plant_index = as.integer(plant_id),
    u0 = plant_intercepts[plant_index],
    height = beta0 +
      beta_temp[as.character(temp)] +
      (beta_week + beta_week_temp[as.character(temp)]) * week +
      u0 +
      rnorm(n(), 0, 1.2)
  ) |>
  select(plant_id, temp, week, height)

# Write out so students can re-load it
# 2027 fix: relative path instead of here()
write_csv(plant_growth, "datasets/plant_growth_repeated.csv")
```

Read in the data:

```{r}
# 2027 fix: read_csv() reads temp as a number, so make it a factor (three temperatures)
plant_growth <- read_csv("datasets/plant_growth_repeated.csv") |>
  mutate(temp = factor(temp, levels = c("10", "15", "20")))
```

<!-- 2027 fix: explain why temp is converted to a factor -->
`read_csv()` reads `temp` as a number (10, 15, 20). Because we treat temperature as a treatment with three levels, we convert it to a factor with `factor()`.


```{r}
head(plant_growth)
```

### Explore the data

First we make a plot of the data:

```{r}
#| echo: false
ggplot(plant_growth, aes(x = week, y = height, group = plant_id, color = temp)) +
  geom_line(alpha = 0.25) +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = "Week", y = "Height (cm)", color = "Temp (°C)")
```

We can see data from each individual plant because it is connected by a line. We also see that plants at higher temperature tend to be taller. And we see that plants differ in their baseline height (week 0).

There are two challenges here:
1. We have repeated measures on the same plants (non-independence).
2. We have variation among plants in baseline height.

A mixed model can handle both of these.

### Wrong model: treat all rows as independent

```{r}
m_lm <- lm(height ~ week * temp, data = plant_growth)
anova(m_lm)
```

This model pretends there are $30 \times 8 = 240$ independent data points.
But most of that information is repeated measures on the same plants.

<!-- 2027 fix: residual df is 234 when temp is a factor (240 - 6 parameters); was 236 for numeric temp -->
We can see that this is pseudoreplication because there are many more residual degrees of freedom (`r df.residual(m_lm)`) than plants (30). This is a classic sign of pseudoreplication. We really cannot trust the p-values or confidence intervals from this model.



### Mixed model: random intercept for plant

Now let's fit a mixed model with plant identity as a random effect:

```{r}
m_lmm1 <- lmer(height ~ week * temp + (1 | plant_id), data = plant_growth)
anova(m_lmm1)
```

In the anova table, note that there are no degrees of freedom and no p-values for fixed effects. This is because calculating these in mixed models is complicated and there are multiple methods. We will look at this more later in this chapter.

Interpretation:

- fixed effects describe the **average** relationship between height, week, and temperature
- the random intercept captures baseline differences among plants


### model checking

Mixed-model diagnostics (i.e., model checking) can be more involved than `lm()`, but you can still start with:

- residual vs fitted plot (nonlinearity / heteroscedasticity)
- normal Q–Q of residuals (approximate)
- check random effect estimates for extreme outliers

```{r}
#| echo: false
par(mfrow = c(2,2))
``` 

1) Residuals vs fitted

```{r}
plot(fitted(m_lmm1), resid(m_lmm1),
     xlab = "Fitted values", ylab = "Residuals")
abline(h = 0, lty = 2)
```

2) Normal Q–Q of residuals

```{r}
qqnorm(resid(m_lmm1)); qqline(resid(m_lmm1))
```


3) Scale–location (sqrt(|resid|) vs fitted)

```{r}
plot(fitted(m_lmm1), sqrt(abs(resid(m_lmm1))),
     xlab = "Fitted values", ylab = "sqrt(|Residuals|)")
```


4) Residuals vs explanatory variable (helpful for nonlinearity with a continuous explanatory variable, here week)

```{r}
# 2027 fix: residuals against the continuous explanatory variable (week), with a proper axis label
plot(plant_growth$week, resid(m_lmm1), xlab = "Week", ylab = "Residuals")
abline(h = 0, lty = 2)
```

```{r}
#| echo: false
par(mfrow = c(1,1))
```



::: {.callout-caution}
**Don’t over-interpret p-values**  
In mixed models, inference depends on how you handle degrees of freedom and uncertainty.
In BIO144, focus on:
- correct model structure (what must be random?)
- effect sizes and uncertainty (CIs)
- sensible plots and biological interpretation
:::


### Extension: random slopes for week

If you believe plants differ in growth rate (not only baseline), add a random slope:

```{r}
m_lmm2 <- lmer(height ~ week * temp + (1 + week | plant_id), data = plant_growth)
summary(m_lmm2)
```

<!-- 2027 fix: explain the singular fit of m_lmm2 -->
Note the message "boundary (singular) fit". In the "Random effects" part of the summary, the standard deviation among plants in their week slope is almost zero (`r round(as.data.frame(VarCorr(m_lmm2))$sdcor[2], 2)`), and the correlation between random intercepts and slopes is `r round(as.data.frame(VarCorr(m_lmm2))$sdcor[3], 2)`. This means the data contain almost no evidence that plants differ in growth rate: the random-slope model is more complex than the data can support (see the caution box about random slopes above). A singular fit is a sign that we should simplify the model, here by going back to the random-intercept model.

Compare the two mixed models:

```{r}
m_lmm_comparison <- anova(m_lmm1, m_lmm2)
m_lmm_comparison
```

<!-- 2027 fix: p-value from inline R (was hard-coded 0.77) -->
There is little evidence that adding random slopes improves the model here (p = `r round(m_lmm_comparison[["Pr(>Chisq)"]][2], 2)`). But in other datasets it might.

::: {.callout-note}
This likelihood ratio test compares nested models. It is widely used, but has subtleties for random effects. In BIO144 you can treat it as a reasonable practical tool, while noting that "testing random effects" is an advanced topic.
:::

### Visualising fitted values

We can visualise the fitted values from both mixed models:

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5
plant_growth_fitted <- plant_growth |>
  mutate(fit_lmm1 = predict(m_lmm1),
         fit_lmm2 = predict(m_lmm2))

p_fit1 <- ggplot(plant_growth_fitted, aes(week, height, group = plant_id, color = temp)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_line(aes(y = fit_lmm1), alpha = 0.35) +
  labs(title = "Random intercept", y = "Height (cm)")

p_fit2 <- ggplot(plant_growth_fitted, aes(week, height, group = plant_id, color = temp)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_line(aes(y = fit_lmm2), alpha = 0.35) +
  labs(title = "Random intercept + slope model", y = "Height (cm)")

p_fit1 + p_fit2
```

There is not much difference between the two models here, but in other datasets random slopes can make a big difference.

### Significance testing for fixed effects

Calculating p-values for fixed effects in mixed models is complicated, and there are multiple methods (Satterthwaite, Kenward-Roger, likelihood ratio tests, bootstrapping, Bayesian credible intervals). It is difficult because the degrees of freedom depend on the random effects structure and the data. There is no clear and objective method to get the degrees of freedom.

<!-- 2027 fix: typos -->
Nevertheless, we can get p-values for terms using the `lmerTest` package (optional). This changes the `lmer()` function to provide p-values using Satterthwaite's method for degrees of freedom.

```{r}
library(lmerTest)
m_lmm1 <- lmer(height ~ week * temp + (1 | plant_id), data = plant_growth)
anova(m_lmm1)
```

<!-- 2027 fix: explain why the Sum Sq differ from the lme4 anova() table above -->
You may notice that the Sum Sq (and F value) for `temp` differ from the `anova()` table we got earlier with `lme4`. This is not because the model changed (it is the same model), but because the two `anova()` functions calculate different types of sums of squares. The `lme4` version gives *sequential* (Type I) sums of squares, like `anova()` for `lm()`: each term is tested after the terms above it. The `lmerTest` version gives Type III sums of squares by default: each term is tested after *all* other terms, including the interaction. The last term (`week:temp`) is the same in both. If you want the sequential version with p-values, use `anova(m_lmm1, type = 1)`. As with `lm()`, when there is an interaction, focus first on the interaction term.

### Reporting (template)

* We modelled plant height as a function of week, temperature, and their interaction using a linear mixed model with plant identity as a random effect.  
* Height increased with week (fixed effect of week), and plants at higher temperature were taller on average (fixed effect of temperature).
* All fixed effects had a p-value < 0.05, calculated using Satterthwaite's method for degrees of freedom (lmerTest package).
* Including plant as a random effect accounted for non-independence due to repeated measures.

## Review

- Mixed models are used when observations are **grouped** (non-independent).
- They combine **fixed effects** (average relationships) with **random effects** (group-to-group variation).
- Random intercepts model different baselines among groups; random slopes allow different responses.
- Nested and crossed random effects reflect study design.
- Mixed models let you keep all observations while avoiding pseudoreplication and false confidence.

<!---
**Think–Pair–Share** (#a_why_random_effects) What problem do random effects solve that fixed effects alone cannot?
--->

<!---
**Think–Pair–Share** (#a_partial_pooling_intuition) How does partial pooling differ from complete pooling or no pooling at all? Hint: pooling is sharing of information.
--->



## Further reading (optional): mixed models

In this course, there is only a short introduction to mixed (multilevel) models. Students who are curious and would like to explore this topic further (purely for their own interest) may find the following resources useful. Material from these resources will not be examined in the final exam, unless it is also already present in the course book.

**McElreath, R. – Statistical Rethinking** This book offers an excellent conceptual introduction to hierarchical (mixed) models, with a strong focus on understanding why they are useful. It introduces ideas such as partial pooling and shrinkage in a very intuitive way. The book uses a Bayesian perspective, but the conceptual insights are valuable even if you later apply frequentist methods. Strongly recommended by Owen for conceptual understanding.

**Gelman, A. & Hill, J. – Data Analysis Using Regression and Multilevel/Hierarchical Models** A classic and very practical text on regression and mixed models, with many applied examples. It complements Statistical Rethinking well and is particularly useful if you want to understand mixed models as they are commonly used in practice.

**R documentation and tutorials for mixed-model packages** If you are interested in implementation in R, the vignettes and documentation for packages such as lme4 (frequentist) or brms (Bayesian) provide hands-on examples of fitting, interpreting, and extending mixed models.

These resources are entirely optional and intended for students who wish to deepen their understanding beyond the scope of the course.
